---
title: "Exploritory Data Analysis (EDA) - Diamonds"
author: "Tim Moore"
date: "February 19, 2017"
output: 
  html_document: 
    css: diamonds.css
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load libraries
library(ggplot2)
library(dplyr)
library(corrplot)

# force output to ignore scientific notation
options(scipen=999)

# load diamonds dataset and remove rows with na data, drop unwanted 
# measurements variable and remove original dataset from memory

load("~/datasets/BigDiamonds.Rda")
clean_diamonds <- filter(subset(diamondsbig, select=-c(measurements)), !is.na(price))
rm(diamondsbig)

#refactor cut, color and clarity to fix legends of charts
clean_diamonds$color <-factor(clean_diamonds$color, 
                              levels=rev(levels(clean_diamonds$color)))

clean_diamonds$clarity <- factor(clean_diamonds$clarity,
                                 levels=rev(levels(clean_diamonds$clarity)))

clean_diamonds$cut <- factor(clean_diamonds$cut, 
                             levels=rev(levels(clean_diamonds$cut)))
```

This report explores a dataset containing price, certification, and 9 physical 
attributes for approximately 597,000 diamonds. The dataset was created by 
Solomon Messing in 2014, and can be found 
[here](https://github.com/SolomonMg/diamonds-data).

##Summary Statistics
###Dimensions of dataset, 597,311 observations over 11 variables.
```{r, echo=FALSE}
dim(clean_diamonds)
```
###Summary of variables
```{r, echo=FALSE}
summary(clean_diamonds)
```
## Univariate plots
```{r plot theme, include=FALSE}
plot_theme <- theme_minimal() +
  theme(plot.title = element_text(size = rel(2), 
                                  color = 'steelblue',
                                  hjust = 0.5), 

                    axis.text = element_text(size = rel(.8))) 


# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

```
###Carat
Interestingly there are peaks that occur at for each integer value up to 7 
carats, similarly visible peaks occur at the .5 carat values from 0.5-3.5 carat
diamonds. This may be due to cultural stigma about purchasing a diamond under a
certain weight, or buyers may prefer to purchase a diamond of lesser cut, color
or clarity that meets or exceeds these carat values.

```{r histogram carat, fig.height = 5,echo=FALSE}
ggplot(clean_diamonds, aes(carat)) + 
  geom_histogram(binwidth= .05) + scale_x_continuous(breaks=0:10) + 
  plot_theme + ggtitle('Histogram by Carat')
```

Carat summary statistics:

```{r, echo=FALSE}
summary(clean_diamonds$carat)
```


###Price

```{r histogram price, fig.height=5, echo=FALSE}
ggplot(clean_diamonds, aes(price)) +
  geom_histogram(bins=1000) + 
  scale_x_continuous(breaks = seq(0,100000, 10000)) + plot_theme +
  labs(title='Histogram by Price', x='Price ($)', y='Count')

```

Transforming from a linear to log distribution of prices to better understand 
the shape of my data. There are two distinct peaks, one around $800, and a 
second peak around $12,500. It appears the market for diamonds is actually two 
separate markets, one for diamonds priced up to ~$12,500, and a second market 
for diamonds priced from $12,500 on up.

```{r histogram log price, fig.height=5, echo=FALSE}
ggplot(clean_diamonds, aes(price)) +
  geom_histogram(bins=1000) + 
  scale_x_log10() + plot_theme +
  labs(title='Histogram by Price', x='log10( Price ($) )', y='Count')
```

Price summary statistics:
```{r, echo=FALSE}
summary(clean_diamonds$price)
```


###Cut, color and clarity

Observe the cut quality of the diamonds is rightward skewed, with most diamonds
having a cut quality of 'ideal'. 

```{r histogram cut color and clarity, fig.height=5, echo=FALSE}
ggplot(clean_diamonds, aes(cut)) + 
  geom_bar(aes(fill=cut)) + plot_theme + labs(title='Bar chart by cut quality',
                                       x='Cut Quality', 
                                       y='Count') 
```

Cut summary statistics:

```{r, echo=FALSE}
summary(clean_diamonds$cut)
```
Similarly, color quality (lower letter is better) is rightward skewed as well. 
It appears most consumers are satisfied with a diamond of color H or better.
```{r bar chart color, fig.height=5, echo=FALSE}
ggplot(clean_diamonds, aes(color)) +
  geom_bar(aes(fill=color)) + plot_theme + labs(title='Bar chart by color',
                                 x='Color',
                                 y='Count')
```

Color summary statistics:
```{r, echo=FALSE}
summary(clean_diamonds$color)
```


Clarity seems to have a threshold of SI2, relatively few diamonds for jewelry 
purposes are sold below this clarity. More than half of all diamonds sold have 
a clarity of at least VS2. Surprisingly more than 5% of all diamonds are 
considered internally Flawless (IF). Additionally more diamonds are classified 
IF than are classified I1 and I2 combined. 

```{r bar chart clarity, fig.height=5, echo=FALSE}
ggplot(clean_diamonds, aes(clarity)) +
  geom_bar(aes(fill=clarity)) + plot_theme + labs(title='Bar chart by Clarity',
                                 x='Clarity',
                                 y='Count')

```

Clarity summary statistics:
```{r, echo=FALSE}
summary(clean_diamonds$clarity)
```


###Certification agency

GIA certifies the vast majority of the diamonds included in this dataset, far 
more than all other certification agencies combined. I wonder if the different 
certification agencies specialize in different types of diamonds. Which agency 
has the highest proportion of low-quality diamonds? Which agency has the 
highest median price? Do you get more diamond for your money from some 
agencies?
```{r bar chart certs, fig.height=5, echo=FALSE}
ggplot(clean_diamonds, aes(cert)) +
  geom_bar(aes(fill=cert)) + plot_theme + 
  theme(axis.text.x = element_text(angle=90, hjust=1)) +
  labs(title='Bar chart by certification agency',
       x = 'Certification Agency',
       y = 'Count')
```

Certification summary statistics:
```{r, echo=FALSE}
summary(clean_diamonds$cert)
```

###Depth
```{r, fig.height=5, echo=FALSE}
ggplot(clean_diamonds, aes(depth)) +
  geom_histogram(bins=150) + plot_theme +
  labs(title='Depth', x='Depth', y='Count')
```

There are a 8066 diamonds with a depth of 0, and another 663 diamonds with a 
depth between 0 and 10 mm. Looking at the summary statistics below makes it 
seem these values may be a data entry error, perhaps the data is off by an 
order of magnitude? 

I find it interesting >12% of all IGI certified diamonds fall into this 
potential error case. I'll pay attention to this certification agency going 
forward to see if any other discrepancies arrise.

```{r, echo=FALSE}
summary(subset(clean_diamonds, depth<=10))

```

Here is a histogram excluding diamonds with a depth of < 10 mm.

```{r, fig.height=5, echo=FALSE}
ggplot(subset(clean_diamonds, depth>10 & !is.na(depth)), aes(depth)) +
  geom_histogram(bins=150) + plot_theme +
  labs(title='Depth', x='Depth', y='Count')
```

Depth summary statistics (>10mm depth):

```{r, echo=FALSE}
summary(subset(clean_diamonds, depth>10)$depth)
```

The sweet spot for diamond depth is betwen 60 mm and 65 mm.

###Table

```{r, echo=FALSE}
ggplot(clean_diamonds, aes(table)) +
  geom_histogram(bins=100) +
  plot_theme +
  labs(title='Table %', x ='Table (%)', y = 'Count')
```

We see a similar issue as above with 2981 diamonds with a table value of 0%,
while another 598 diamonds have a table value between 0% and 10%. Looking 
deeper at these data points reveal all other variables are populated with the 
exception of depth. Again IGI appears to be represented at a higher rate than 
would be expected. 


Below we see the summary stats for diamonds with a table of < 10%:
```{r, echo=FALSE}
summary(subset(clean_diamonds, table <10))
```

Table summary statistics (>10% table):

```{r, echo=FALSE}
summary(subset(clean_diamonds, table > 10)$table)
```
The sweet spot for table is between 55-60%.

```{r, echo=FALSE}
ggplot(subset(clean_diamonds, table >10), aes(table)) +
  geom_histogram(bins=100) +
  plot_theme +
  labs(title='Table %', x ='Table (%)', y = 'Count')
```


###X, Y, and Z measurements
```{r, echo=FALSE}
ggplot(subset(clean_diamonds, !is.na(x)), aes(x)) +
  geom_histogram(bins=1000) +
  plot_theme +
  labs(title='X (mm)', x='X (mm)', y='Count')


```

1,814 diamonds have a n/a value for x-axis measurement otherwise they seem to
be mostly complete entries. Below you will find the summary statistics for
these diamonds.

```{r, echo=FALSE}
summary(subset(clean_diamonds, is.na(x)))
```

```{r, echo=FALSE}
ggplot(subset(clean_diamonds, !is.na(y)), aes(y)) +
  geom_histogram(bins=1000) +
  plot_theme +
  labs(title='Y (mm)', x='Y (mm)', y='Count')
```

1851 diamonds contain n/a values for y-axis measurement. Below are the summary
statistics for these diamonds.

```{r, echo=FALSE}
summary(subset(clean_diamonds, is.na(y)))
```

```{r, echo=FALSE}
ggplot(subset(clean_diamonds, !is.na(z)), aes(z)) +
  geom_histogram(bins=1000) +
  plot_theme +
  labs(title='Z (mm)', x='Z (mm)', y='Count')
```

2543 diamonds contain n/a values for the z-axis measurement. Below are the
summary statistics for these diamonds.

```{r, echo=FALSE}
summary(subset(clean_diamonds, is.na(z)))
```

###Cleaning the dataset

Given the size of this dataset, I feel comfortable dropping any diamonds with 
a x, y, z, table or depth value less than 10 (% or mm respectively) or n/a. I
will be using this revised dataset for the remainder of this analysis. This
leaves me with 585,808 diamonds to examine. I've saved the 'dirty' dataset of
12,223 diamonds to analyze in more depth below.

Summary statistics for this cleaned dataset are below:

```{r, echo=FALSE}
cleaned <-subset(clean_diamonds, 
                 !is.na(x) & !is.na(y) & !is.na(z) & table>=10 & depth >=10)

dirty <- subset(clean_diamonds, 
                is.na(x) | is.na(y) | is.na(z) | table<10 | depth <10)

summary(cleaned)
```

###Best vs. worst quality diamonds

Looking below at diamonds that have ideal cut, color of D, and clarity of IF.
```{r, echo=FALSE}
ggplot(subset(cleaned, cut == 'Ideal' & color == 'D' & clarity == 'IF'),
       aes(price)) +
  geom_histogram(bins=1000) +
  plot_theme +
  labs(title='Prices of Best-quality Diamonds', 
       x='Price ($)', 
       y='Count')
```

This plot is hard to see much of what is going on, I've replotted it below 
using a base 10 log scale for price. 
```{r, echo=FALSE}
ggplot(subset(cleaned, cut == 'Ideal' & color == 'D' & clarity == 'IF'),
       aes(price)) +
  geom_histogram(bins=1000) +
  plot_theme + scale_x_log10() +
  labs(title='Log Prices of Best-quality Diamonds', 
       x='Log_10 (Price ($))', 
       y='Count')
```

Here are the summary statistics for the best quality diamonds. The most 
facinating bit to me is the max size is 2.58 carat vs 9.25 carats for the 
heaviest stone in the complete dataset.

```{r, echo=FALSE}
summary(subset(cleaned, cut == 'Ideal' & color == 'D' & clarity == 'IF'))
```


Now let's check out the other end of the spectrum, what do the lowest quality 
diamonds look like?

```{r, echo=FALSE}
ggplot(subset(cleaned, cut =='Good' & color == 'L' & clarity == 'I2'), 
       aes(price)) +
  geom_histogram(bins= 50) +
  plot_theme +
  labs(title='Prices of Worst-quality diamonds', x='Price ($)', y='Count')

```

I'm intrigued the median and mean carat values for the worst quality diamonds 
match up almost exactly with the best quality diamonds. Both right around 1 
carat!

```{r}
summary(subset(cleaned, cut =='Good' & color == 'L' & clarity == 'I2'))
```

##Dataset
###Structure of dataset
After cleaning my dataset of non-sensical and n/a values, 585,088 diamonds 
remain with 11 features (carat, cut, color, clarity, table, depth, 
certification agency, price, x, y, and z). The variables cut, color and clarity
are ordered factor variables with the following levels:

(worst) ------------> (best)

Cut: Good, Very Good, Ideal

Color: L, K, J, I, H, G, F, E, D

Clarity: I3, I2, I1, SI2, SI1, VS2, VS1, VVS2, VVS1, IF

other observations:

- More diamonds are of Ideal cut than are of the other two cuts combined.

- Median carat size is 0.900.

- Most diamonds are color G or above.

- 75% of all diamonds in my cleaned dataset are 1.5 carat or less.

- Median price is $3,539 with a high value of $99,990.

###Features of interest
The features that are most interesting as output values for a model are carat
and price. I'm interested in diving into the differences between the different
certification agencies. Do some agencies specialize in lesser-quality diamonds?
Additionally I'm excited to look at the spike in the number of diamonds sold 
with weights at or above integer valued carats. I am intrigued to see if lesser
cut, clarity or color diamonds are kept bigger to sell at or above these 
integer values.

##Bivariate and Multivariate plots
###Correlation matrix

```{r echo=FALSE}

cleaned_cor <- cleaned[c('price', 'carat', 'table', 'depth', 'x', 'y', 'z')]
cor_matrix <- cor(cleaned_cor)
round(cor_matrix, 2)
corrplot(cor_matrix, method='number', outline=TRUE, tl.srt=45)
```

###Price vs. Carat
The following plots examine the relationship between price and carat weight
for diamonds. First we will look at the overall density plot. Note the large 
vertical streaks that occur at carat weights ending with .50 and .00. I will be
referring to this preference for diamonds to exceed a 'round' carat weight as 
a vanity metric. 

```{r echo=FALSE, fig.height=8}
ggplot(cleaned, aes(carat, price)) +
  geom_point(alpha = .01, shape = 1) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  labs(title='Price vs. Carat', 
       x='Carat', y='Price') +
  guides(colour = guide_legend(override.aes = list(alpha=.5)))
```

###Price vs. Carat - by Cert. Agency
Let's start by looking at price vs carat by certification agency to see if any 
agencies have specialties. GIA dominates the certification market in sheer 
nubers as well as price. EGL USA and IGI both bring less of a price premium on
diamonds they certify.
```{r, echo=FALSE, fig.height=8}
ggplot(cleaned, aes(carat, price, color = cert)) +
  geom_point(alpha = .01, shape = 1) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(title='Price vs. Carat - by Cert. Agency', 
       x='Carat', y='Price') 

```

###Price vs. Carat - Color
Looking at the the same plot colored by the diamond's color. It does appear a 
lot of 2 & 3 carat diamonds of lesser color were allowed to come to market 
relative to other weights.

```{r, echo=FALSE, fig.height=8}
ggplot(cleaned, aes(carat, price, color = color)) +
  geom_point(alpha = .01, shape = 1) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  labs(title='Price vs. Carat - by Diamond Color', 
       x='Carat', y='Price') 

```

###Price vs. Carat - Cut
Now looking at the smae plot by cut quality, not many non-ideal cuts are allowed
to come to market.

```{r, echo=FALSE, fig.height=8}
ggplot(cleaned, aes(carat, price, color = cut)) +
  geom_point(alpha = .01, shape = 1) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  labs(title='Price vs. Carat - by Cut', 
       x='Carat', y='Price')
```

###Price vs. Carat - Clarity
Similar to color, we observe more diamonds of lesser quality to come to market 
at 1, 1.5, 2 and 3 carat weights. 

```{r, echo=FALSE, fig.height=8}
ggplot(cleaned, aes(carat, price, color = clarity)) +
  geom_point(alpha = .01, shape = 1) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  labs(title='Price vs. Carat - by Clarity', 
       x='Carat', y='Price') +
  guides(colour = guide_legend(override.aes = list(alpha=1)))
```

###Quality vs price - Vanity weights
Now, I'm getting really intersted in the behaviour of diamond prices vs. 
quality at the vanity points. Below we can start to observe how prices change
with differing color and/or clarity at the vanity points. Additionally notice
the how few diamonds are offered in the low quality quadrant, perhaps diamonds 
with these ratings are used industrially instead of for jewelry?

```{r echo=FALSE,fig.height=12, fig.width=12}
two_carat_price <- ggplot(subset(
  cleaned, carat %in% c(2.00, 2.01, 2.02, 2.03, 2.04, 2.05)), 
  aes(clarity, color, color = price)) +
  geom_jitter(alpha=.15) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_gradientn(colours = terrain.colors((10)), guide='colourbar') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(title='Quality vs Price for 2 Ct Diamonds', x='Clarity', y='Color')

one_carat_price <-ggplot(subset(
  cleaned, carat %in% c(1.00, 1.01, 1.02, 1.03, 1.04, 1.05)), 
  aes(clarity, color, color = price)) +
  geom_jitter(alpha=.15) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_gradientn(colours = terrain.colors((10)), guide='colourbar') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(title='Quality vs Price for 1 Ct Diamonds', x='Clarity', y='Color')

one_point_five_price <- ggplot(subset(
  cleaned, carat %in% c(1.50, 1.51, 1.52, 1.53, 1.54, 1.55)), 
  aes(clarity, color, color = price)) +
  geom_jitter(alpha=.15) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_gradientn(colours = terrain.colors((10)), guide='colourbar') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(title='Quality vs Price for 1.5 Ct Diamonds', x='Clarity', y='Color')

two_point_five_price  <- ggplot(subset(
  cleaned, carat %in% c(2.50, 2.51, 2.52, 2.53, 2.54, 2.55)), 
  aes(clarity, color, color = price)) +
  geom_jitter(alpha=.15) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_gradientn(colours = terrain.colors((10)), guide='colourbar') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(title='Quality vs Price for 2.5 Ct Diamonds', x='Clarity', y='Color')

multiplot(one_carat_price,two_carat_price, 
          one_point_five_price, two_point_five_price, cols=2)
```

###Quality vs. Cut - Vanity weights
Looking at Diamond Quality and cut quality at the vanity points. There are
thresholds easily visible (I1 clarity and K Color) demonstrating benchmarks to 
exceed for a diamond to be viable on the market.

```{r echo=FALSE, fig.height=12, fig.width=12}
two_carat_cut <- ggplot(subset(
  cleaned, carat %in% c(2.00, 2.01, 2.02, 2.03, 2.04, 2.05)), 
  aes(clarity, color, color = cut)) +
  geom_jitter(alpha=.05) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_brewer(palette = 'Set1') +
  guides(colour = guide_legend(override.aes = list(alpha=1)), reverse=TRUE) +
  labs(title='Quality vs Cut for 2 Ct Diamonds', x='Clarity', y='Color')

one_carat_cut <-ggplot(subset(
  cleaned, carat %in% c(1.00, 1.01, 1.02, 1.03, 1.04, 1.05)), 
  aes(clarity, color, color = cut)) +
  geom_jitter(alpha=.05) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_brewer(palette = 'Set1') +
  guides(colour = guide_legend(override.aes = list(alpha=1)), reverse=TRUE) +
  labs(title='Quality vs Cut for 1 Ct Diamonds', x='Clarity', y='Color')

one_point_five_cut <- ggplot(subset(
  cleaned, carat %in% c(1.50, 1.51, 1.52, 1.53, 1.54, 1.55)), 
  aes(clarity, color, color = cut)) +
  geom_jitter(alpha=.05) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_brewer(palette = 'Set1') +
  guides(colour = guide_legend(override.aes = list(alpha=1)), reverse=TRUE) +
  labs(title='Quality vs Cut for 1.5 Ct Diamonds', x='Clarity', y='Color')

two_point_five_cut  <- ggplot(subset(
  cleaned, carat %in% c(2.50, 2.51, 2.52, 2.53, 2.54, 2.55)), 
  aes(clarity, color, color = cut)) +
  geom_jitter(alpha=.05) + plot_theme + scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_brewer(palette = 'Set1') +
  guides(colour = guide_legend(override.aes = list(alpha=1)), reverse=TRUE) +
  labs(title='Quality vs Cut for 2.5 Ct Diamonds', x='Clarity', y='Color')

multiplot(one_carat_cut,two_carat_cut, 
          one_point_five_cut, two_point_five_cut, cols=2)
```

```{r, echo=FALSE}
vanity <- c(.45,.46,.47,.48,.49,.5,.51,.52,.53,.54,.55,
95,.96,.97,.98,.99,1,1.01,1.02,1.03,1.04,1.05,1.45,1.46,1.47,1.48,1.49,1.5,
1.51,1.52,1.53,1.54,1.55,1.95,1.96,1.97,1.98,1.99,2,2.01,2.02,2.03,2.04,2.05,
2.45,2.46,2.47,2.48,2.49,2.5,2.51,2.52,2.53,2.54,2.55,2.95,2.96,2.97,2.98,2.99,
3,3.01,3.02,3.03,3.04,3.05,3.45,3.46,3.47,3.48,3.49,3.5,3.51,3.52,3.53,3.54,3.55
,3.95,3.96,3.97,3.98,3.99,4,4.01,4.02,4.03,4.04,4.05,4.45,4.46,4.47,4.48,4.49,
4.5,4.51,4.52,4.53,4.54,4.55,4.95,4.96,4.97,4.98,4.99,5,5.01,5.02,5.03,5.04,5.05
,5.45,5.46,5.47,5.48,5.49,5.5,5.51,5.52,5.53,5.54,5.55,5.95,5.96,5.97,5.98,5.99
,6,6.01,6.02,6.03,6.04,6.05,6.45,6.46,6.47,6.48,6.49,6.5,6.51,6.52,6.53,6.54,6.55
,6.95,6.96,6.97,6.98,6.99,7,7.01,7.02,7.03,7.04,7.05,7.45,7.46,7.47,7.48,7.49,
7.5,7.51,7.52,7.53,7.54,7.55,7.95,7.96,7.97,7.98,7.99,8,8.01,8.02,8.03,8.04,8.05
,8.45,8.46,8.47,8.48,8.49,8.5,8.51,8.52,8.53,8.54,8.55,8.95,8.96,8.97,8.98,8.99
,9,9.01,9.02,9.03,9.04,9.05)

cleaned$vain <- ifelse(cleaned$carat %in% vanity, c('Within .05 carat of a round number'), c('Not within .05 carat of a round number'))
```

###Diamond quality and vanity
To dive deeper into the vanity idea, I've replotted all diamonds by color and 
clarity. This time the colors signify if the diamond belongs in the vanity 
group or not. We see top quality diamonds tend not to be vanity diamonds, 
instead the vanity diamonds tend to have a color of J or better, and a clarity 
of SI2-VS1.

```{r echo=FALSE, fig.height=8, fig.width=12}
ggplot(cleaned, aes(clarity, color, color = vain)) +
  geom_jitter(alpha=.05) + plot_theme + 
  scale_x_discrete() + 
  scale_y_discrete(limits=rev(levels(cleaned$color))) +
  scale_color_manual(values=c('#ef8a62', '#67a9cf')) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(title='Diamond Quality and vanity', x='Clarity', y='Color')
```

###Price vs Carat - Subset by Clarity
Observe the price jumps for vanity diamonds, these jumps are consistent across
various clarity levels.

```{r echo=FALSE, fig.height = 12, fig.width=12}
ggplot(cleaned, aes(carat, price, color=vain)) +
  geom_point(alpha = .08, shape = 1) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  facet_grid(clarity~.) +
  labs(title='Price by Clarity', x='Carat', y='Price ($)')
```

###Price vs. Carat - Subset by Color
We see the same trend for a price hike for vanity diamonds across color levels
also

```{r echo=FALSE, fig.height=12, fig.width=12}
ggplot(cleaned, aes(carat, price, color=vain)) +
  geom_point(alpha = .08, shape = 1) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  facet_grid(color~.) +
  labs(title='Price by Color', x='Carat', y='Price ($)')
```

###Price vs. Color - Vanity weights by Cert. Agency
Checking out the color offerings for vanity weight diamonds by certification 
agency. Again we see GIA seems to certify the vast majority of high quality 
diamonds while others specialize in bringing lesser quality diamonds to market.

```{r echo=FALSE, fig.height=12, fig.width=12}
ggplot(subset(cleaned, vain == 'Within .05 carat of a round number'), 
       aes(carat, price, color = color)) +
  geom_point(alpha=.01) + plot_theme + 
  scale_x_continuous(breaks = 0:10) + 
  scale_y_continuous(breaks = seq(0,100000,10000)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  facet_grid(cert~.) +
  labs(title='Price vs Color - near to half carat', 
       x='Carat', y='Price ($)')
```

###Price vs Carat - Subset by Cert Agency (Non-Vanity weights)
Now we are looking at these same plots but for the non-vanity weight diamonds.
GIA diamonds tend to be more expensive as well as carry nearly all of the high-
end market.

```{r echo=FALSE, fig.height=12, fig.width=12}
ggplot(subset(cleaned, vain == 'Not within .05 carat of a round number'), 
       aes(carat, price, color = color)) +
  geom_point(alpha=.01) + plot_theme + 
  scale_x_continuous(breaks = 0:10) + 
  scale_y_continuous(breaks = seq(0,100000,10000)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  facet_grid(cert~.) + 
  labs(title='Price vs Cert. Agency', x='Carat', y='Price ($)')
```

###Examining the Vanity Price jump
Below we can examine the vanity price jump subset by cut, color, clarity and 
certification agency. We observe GIA commands the largest price premium, and 
has a stranglehold on the high-end market (by quality and price).

```{r, echo=FALSE, fig.height=12, fig.width=12}


d_1 <- ggplot(subset(clean_diamonds, carat <=1.1 & carat >=.9), 
              aes( carat, price)) + 
  geom_point(aes(color = clarity), alpha = .15) + 
  plot_theme +
  labs(title='0.9 - 1.1 carat diamonds', y='Price ($)', x = 'Carat Weight') + 
  guides(colour = guide_legend(override.aes = list(alpha=1)))
  

p1 <- d_1 + facet_grid(cert~.)

d_2 <- ggplot(subset(clean_diamonds, carat <=2.1 & carat >=1.9), 
              aes(carat, price)) +
  geom_point(aes(color = clarity), alpha = .15) + 
  plot_theme + 
  labs(title='1.9-2.1 carat diamonds', y='Price ($)', x = 'Carat Weight')+ 
  guides(colour = guide_legend(override.aes = list(alpha=1)))

p2 <- d_2 + facet_grid(cert~.)

d_3 <- ggplot(subset(clean_diamonds, carat <=1.1 & carat >=.9), 
              aes( carat, price)) + 
  geom_point(aes(color = cut), alpha = .15) + 
  plot_theme +
  labs(title='0.9 - 1.1 carat diamonds', y='Price ($)', x = 'Carat Weight') + 
  guides(colour = guide_legend(override.aes = list(alpha=1)))

p3 <- d_3 + facet_grid(cert~.)

d_4 <- ggplot(subset(clean_diamonds, carat <=2.1 & carat >=1.9), 
              aes(carat, price)) +
  geom_point(aes(color = cut), alpha = .15) + 
  plot_theme + 
  labs(title='1.9-2.1 carat diamonds', y='Price ($)', x = 'Carat Weight') + 
  guides(colour = guide_legend(override.aes = list(alpha=1))) 


p4 <- d_4 + facet_grid(cert~.)

d_5 <- ggplot(subset(clean_diamonds, carat <=1.1 & carat >=0.9),
              aes(carat, price)) +
  geom_point(aes(color = color), alpha=.15)+
  plot_theme +
  labs(title='0.9-1.1 carat diamonds', y='Price ($)', x = 'Carat Weight') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) 

p5 <- d_5 + facet_grid(cert~.)

d_6 <- ggplot(subset(clean_diamonds, carat <=2.1 & carat >=1.9),
              aes(carat, price)) +
  geom_point(aes(color = color), alpha=.15)+
  plot_theme +
  labs(title='1.9-2.1 carat diamonds', y='Price ($)', x = 'Carat Weight') +
  guides(colour = guide_legend(override.aes = list(alpha=1))) 

p6 <- d_6 + facet_grid(cert~.)

multiplot(d_1, d_5, d_2, d_6, cols=2)

```


##FInal Plots
###Plot One

```{r, echo=FALSE, fig.height=12, fig.width=12}
ggplot(cleaned, aes(carat, price, color=vain)) +
  geom_point(alpha = .08, shape = 1) +
  plot_theme + scale_x_continuous(breaks=0:10) + 
  scale_y_continuous(breaks=seq(0,100000,10000)) +
  guides(colour = guide_legend(override.aes = list(alpha=1))) +
  labs(title='Vanity price jump', x='Carat', y='Price ($)')
```

###Description One
This plot highlights the price jump that occurs every 0.50 carat. This price 
jump is visible through all of the 3 C's (clarity, cut and color) of a diamond.

###Plot 2

```{r echo=FALSE, fig.height=12, fig.width=18}
carat.class <- cut(cleaned$carat, seq(0, 3.5, .1))
color <- ggplot(cleaned, aes(carat.class)) +
  geom_bar(aes(fill=color), position='dodge') + plot_theme +
  labs(title='Color composition by carat weight',
       x='Carat Weight', y='# of Diamonds')

cut <- ggplot(cleaned, aes(carat.class)) +
  geom_bar(aes(fill=cut), position='dodge') + plot_theme +
  labs(title='Cut composition by Carat Weight', 
       x='Carat Weight', y='# of Diamonds')

clarity <- ggplot(cleaned, aes(carat.class)) +
  geom_bar(aes(fill=clarity), position='dodge') + plot_theme +
  labs(title='Clarity composition by Carat Weight', 
       x='Carat Weight', y='# of Diamonds')

multiplot(color, clarity, cut)

```




###Description 2
This plot makes clear two interesting trends in the data - the spike in quantity
of diamonds sold at vanity points as well as the drop in quality for diamonds at
these same weights. The diamonds have been binned into 0.1 carat buckets. The 
color, clarity and cut proportions were then plotted for each of these bins. 
Observe the spikes in the total # of diamonds sold at the 1.0, 1.5, 2.0 and 3.0 
carat weights. Notice also the disproportionate percentage of low quality 
(holds true for color, clarity & cut) diamonds present at these 4 weights. We 
can see the saturation of the diamond market at 'vanity' weights with lower 
quality diamonds relative to the non-vanity weights.

###Plot 3

```{r echo=FALSE, fig.height=8, fig.width=18}
vain <- subset(cleaned, carat >=.9 & carat <=1.1)
Weight <- cut(vain$carat, 2, labels = c('0.90 - 0.99 ct', '1.00 - 1.09 ct'))

bp1 <- ggplot(vain, aes(x=clarity, y=price, fill=Weight)) + 
  geom_boxplot() + plot_theme + 
  scale_y_continuous(breaks=seq(0,35000, 1000), limits=c(0,35000)) +
  labs(title='Vanity Price Premium - Clarity', x = 'Clarity', y='Price ($)')

bp2 <- ggplot(vain, aes(x=color, y=price, fill=Weight)) +
  geom_boxplot() + plot_theme + 
  scale_y_continuous(breaks=seq(0, 35000, 1000), limits=c(0,35000)) +
  labs(title='Vanity Price Premium - Color', x='Color', y='Price ($)')

bp3 <- ggplot(vain, aes(x=carat, y=price, fill=Weight)) +
  geom_boxplot() + plot_theme + 
  scale_y_continuous(breaks=seq(0, 35000, 1000), limits=c(0,35000)) +
  labs(title='Vanity Price Premium', x='All Diamonds', y='Price ($)')

multiplot(bp3, bp1, bp2, cols = 3)

```

###Description 3

Diving in deeper to the relationship examined in plot 1, I decided to identify
the price premium for a diamond that weighs at least 1 carat. These three box 
plots examine the price premium commanded by diamonds around the
1.00 carat vanity point. First looking at all diamonds, observe the median 
diamond exhibits a price premium of $1,491 once carat weight exceeds 1.00. Next,
notice the price premium broken down by quality of diamond. I find fascinating 
how the median price of a vanity diamond exceeds the upper range of the 
corresponding box for the non-vain diamonds. This relationship holds for all 
diamonds in this weight range with the exception of those with a color rating 
of G. 

These price premiums are surprisingly durable to changes in quality, and 
for the highest quality diamonds can nearly double the price of a corresponding 
diamond of similar size and weight. Looking at median prices, you have to drop
3 levels for either clarity or color to find a vain diamond of similar price.

##Reflection

Starting from a data set of more than 597,000 diamonds across 11 variables from
2014, after cleaning up and removing incomplete entries I was left with just 
over 585,000 diamonds to work with. I began by exploring the individual 
variables, to get a feel for the shape of my data. Next I moved on to explore 
questions of interest as I examined what determines the price of a diamond. 
Eventually I dove deep into the price jump and quality decrease found in 
diamonds just above every 0.50 carat increment. 

There is an observable trend in diamonds of lesser cuts and colors on the market
weighing in at or above x.00 and x.50 carat weight. This trend holds across 
differing certification agencies and persists to 2.5 carat weights.

Also I found prices increase dramatically for 'perfect' (Internally Flawless 
clarity as well as D color) holding carat weight constant. Consumers will 
easily pay 3x as much for a IF/D diamond than they will for a SI2/J (threshold
level below which the quantity available is significantly diminished).

I think it would be quite interesting to investigate the differences in 
Certification agencies diamond portfolios going forward. Especially if I were to
get a location of origin for each diamond. This would open doors to identifying
potential sources of fraud for diamonds coming out of war torn areas, I would 
like to see if certain agencies have a 'blind eye' policy towards location of 
origin. 
